\documentclass{beamer}
\usetheme[pageofpages=of,% String used between the current page and the
                         % total page count.
          bullet=circle,% Use circles instead of squares for bullets.
          titleline=true,% Show a line below the frame title.
          alternativetitlepage=true,% Use the fancy title page.
       %   titlepagelogo=logo-polito,% Logo for the first page.
       %   watermark=watermark-polito,% Watermark used in every page.
       %   watermarkheight=100px,% Height of the watermark.
       %   watermarkheightmult=4,% The watermark image is 4 times bigger
                                % than watermarkheight.
          ]{Torino}

\setbeamertemplate{footline}{
  \begin{beamercolorbox}[wd=\paperwidth,ht=1ex,dp=1ex]{footline}
    \vspace{5pt} \hspace{1em} \insertframenumber/\inserttotalframenumber
  \end{beamercolorbox}
}

\newcommand{\bx}{\boldsymbol{x}}
\newcommand{\given}{\,|\,}

\author{Brendon J. Brewer}
\title{Diffusive Nested Sampling: Not Dead Yet}
\institute{The University of Auckland}
\date{}


\linespread{1.3}
\usepackage{minted}
\usepackage[utf8]{inputenc}
\usepackage{dsfont}


\begin{document}

\frame{\titlepage}


\begin{frame}
\frametitle{Minimalist Slides}
These minimalist slides are inspired by Paul Murrell.

\end{frame}

\begin{frame}
\frametitle{Bayesian Inference}

\begin{itemize}
\item Uncertainty is described using probability theory.\pause
\item Usually you want to calculate posterior distributions for unknown
quantities, given the data you have.\pause
\item Computationally, this is best achieved by generating samples from
the posterior distributions, using methods such as Markov Chain Monte Carlo.
\end{itemize}

\end{frame}



\begin{frame}
\frametitle{Background}

\begin{itemize}
\item Probability theory is used to describe uncertainty.\pause
\item Prior distributions get updated to posterior distributions when you
obtain data. Posterior distributions are the main
conclusion.\pause
\item The marginal likelihood is a single number that is also useful
for model comparison purposes. Think: something like AIC but philosophically
better (though, not idiot proof).
\end{itemize}

\end{frame}


\begin{frame}
\frametitle{Computing the Results}
\begin{itemize}
\item The best way to describe high dimensional probability distributions
is with samples.\pause
\item Markov Chain Monte Carlo (MCMC) methods `explore the parameter space'
to generate samples.\pause
\item Most cannot calculate marginal likelihoods, or do so poorly.\pause
\item Has failure modes similar to optimisation techniques, e.g., when
the posterior is multimodal.
\end{itemize}


\end{frame}




\begin{frame}
\frametitle{Two Categories of MCMC Methods}
\begin{itemize}
\item Different ways of moving around. Metropolis, Gibbs, Slice, Hamiltonian,
Affine Invariant, etc.\pause
\item Change the distribution. Annealing, etc.
\end{itemize}


\end{frame}



\begin{frame}
\frametitle{Nested Sampling}
\begin{itemize}
\item Invented by physicist John Skilling in 2004.\pause
\item The gold standard for calculating marginal likelihood (a high dimensional
integral), and also gives posterior samples.\pause
\item Starts with points representing the prior distribution (i.e., guesses),
moves them `uphill' in likelihood, at a controlled rate. Like an optimiser
with a speedometer and brakes.
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Diffusive Nested Sampling}
\begin{itemize}
\item Invented at the MaxEnt 2009 conference in Oxford, Mississippi.\pause
\item A `pure MCMC' version of Nested Sampling.\pause
\item Replace posterior distribution with `mixture of constrained priors'.\pause
\item Easier to explore. Can jump between modes. Gives marginal likelihood.
Posterior available by reweighting.\pause
\item Canonical implementation is DNest4, written in 2018. It's in C++,
with some Python postprocessing scripts.
\end{itemize}
\end{frame}


\begin{frame}
\frametitle{DNest4 Demos}
\begin{itemize}
\item `Road' dataset (simple linear regression).\pause
\item Implementation as a C++ class (runs fast but a lot of work
compared to \mintinline{r}{lm()} --- even loading
data requires a bunch of I/O programming).\pause
\item The perturb function is also a lot of work.\pause
\item Galaxy demo (gaussian mixture model).
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{DNest4 Users}
\begin{itemize}
\item I don't have many users in the wild --- the learning curve is steep.
People prefer bad samplers that are easy to use.\pause
\item Needed ways of writing models simply in easier languages than C++.\pause
\item Python, R, and Julia now supported.\pause
\item You need to write a function to define the prior and a function
to define the log likelihood.
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{DNest4 Demos}
\begin{itemize}
\item Road example implemented in R.\pause
\item The R connection is by Simon Urbanek.\pause
\item Python and Julia examples.\pause
\item The Python and Julia connections were by myself and ChatGPT.
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Strengths and Weaknesses}
\begin{itemize}
\item Strengths:
works on hard problems with multimodality, strong correlations, phase
transitions, etc, giving posterior samples and marginal likelihood.\pause
\item Weaknesses: Tends to only be sane on problems where likelihood evaluation
is pretty fast. Avoid on problems with very tiny (relative to prior volume)
yet simple posteriors.
\end{itemize}

\end{frame}


\end{document}

